---
title: "DataVisualization_ElwhaRiver"
author: "Kathleen Mason"
date: "4/20/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r Getting Setup}
getwd()

library(tidyverse)
library(viridis)
library(ggplot2)

USGS_ElwhaRiverDischarge<-read_csv("./Processed/USGS_ElwhaRiver_processed.csv")

class(USGS_ElwhaRiverDischarge$Date) #Still a date, that's good


```

```{r Theme}

mytheme <- theme_light(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)


#This should add a caption?
#+labs(subtitle = "theme_gray()")
```


##Visualization Plot Assignment
```{r Useful Visualization}

#lines to describe what is happening
d=data.frame(date=as.Date(c("2011-09-15", "2014-08-26")), event=c("Dam Removal Project Begins", "Dam Removal Project Complete"))



myplot<-ggplot(USGS_ElwhaRiverDischarge)+
  geom_line(aes(x = Date, y = DailyDischarge), size=.4) +
geom_vline(data=d, mapping=aes(xintercept=date), color="skyblue4", lty=5, size=.5)+
  geom_hline(yintercept = 47.7, lty=3, color="sienna3", size=.7)+
  geom_text(data=d, mapping=aes
            (x=date, y=200, label=event), 
            size=3, angle=90, vjust=-0.4, hjust=0, color= "skyblue4", fontface="bold")+
  geom_text(x= as.Date("2015-11-01"), y=20, size = 3, label= "Mean Daily Discharge", color= "sienna3", vjust=-0.4, hjust=0, fontface="bold")+
  mytheme+
  ylab(expression(Daily~Discharge~(m^2/s)))+
  xlab("Date")+
  theme(axis.title.y = element_text(size = rel(.75), angle = 90))+
  theme(axis.title.x = element_text(size= rel(.75)))+
  labs(caption ="Figure 1. Daily water discharge (m^2/s) from the Elwha River from September 15, 2011 to September 30, 2016 measured at the \nU.S. Geological Survey gaging station 12046260 at the diversion near Port Angeles, Washington. A project to remove \nthe Elwha and Glines Canyon Dam began on September 15, 2011, and was completed on August 26, 2014. Mean Daily \ndischarge across the whole time range was 47.7.")+
  theme(plot.caption = element_text(hjust = 0) )
  print(myplot)

#good as is but doesn't have superscript in the caption


  ggsave("./Output/USGS_ElwhaDailyDischarge_plot.jpg", plot=myplot, width =10, height=6, units="in", dpi=300)
  
  #Now for total sediment dischagre
  d=data.frame(date=as.Date(c("2011-09-15", "2014-08-26")), event=c("Dam Removal Project Begins", "Dam Removal Project Complete"))



myplot<-ggplot(USGS_ElwhaRiverDischarge)+
  geom_line(aes(x = Date, y = TotalSedimentDischarge), size=.4) +
geom_vline(data=d, mapping=aes(xintercept=date), color="skyblue4", lty=5, size=.5)+
  geom_hline(yintercept = 9886.876, lty=3, color="sienna3", size=.7)+
  geom_text(data=d, mapping=aes
            (x=date, y=100000, label=event), 
            size=2, angle=90, vjust=-0.4, hjust=0, color= "skyblue4", fontface="bold")+
  geom_text(x= as.Date("2016-05-01"), y=20, size = 2, label= "Mean Total \nSediment \nDischarge", color= "sienna3", vjust=-0.4, hjust=0, fontface="bold")+
  mytheme+
  ylab(expression("Total Daily Sediment Discharge (tonnes)")+
  xlab("Date")+
  theme(axis.title.y = element_text(size = rel(.75), angle = 90))+
  theme(axis.title.x = element_text(size= rel(.75)))
  print(myplot)

  #fig captions=Daily total sediment discharge from the Elwha River from September 15, 2011 to September 30, 2016 measured at the U.S. Geological Survey gaging station 12046260 at the diversion near Port Angeles, Washington. A project to remove the Elwha and Glines Canyon Dam began on September 15, 2011, and was completed on August 26, 2014.Mean Daily sediment discharge across the whole time range was 9886.876 tonnes.
```


##Analysis
Question: How does water and sediment discahrge change during and after the dam removal process? >Is sediment discharge different between during and after dam removal? Is water discharge different between during and after dam removal?
>Need to divide the explanatory value of time into two categories- during dam removal (2011-09-15 to 2014-08-26) and after dam removal (2014-08-27 to 2016-09-30). The response variable is discharge. Analysis will be performed using a two sample ttest. One ttest on total sediment discharge over time, and one on total discharge (water) over time. 

```{r Creating during and post dam removal data}
USGS_ElwhaRiverDischarge_During <- USGS_ElwhaRiverDischarge %>% filter(Date < as.Date("2014-08-26"))

USGS_ElwhaRiverDischarge_After <- USGS_ElwhaRiverDischarge %>% filter(Date > as.Date("2014-08-26"))

#Add "after" and "during" columns
USGS_ElwhaRiverDischarge_After$TimeStamp<-"After"
USGS_ElwhaRiverDischarge_During$TimeStamp<-"During"

USGS_ElwhaRiverDischarge_TimeStamp<- full_join(USGS_ElwhaRiverDischarge_After, USGS_ElwhaRiverDischarge_During)

write.csv(USGS_ElwhaRiverDischarge_TimeStamp, row.names = FALSE, file =("./Processed/USGS_ElwhaRiverDischarge_TimeStamped_processed.csv"))
write.csv(USGS_ElwhaRiverDischarge_After, row.names = FALSE, file =("./Processed/USGS_ElwhaRiverDischarge_After_processed.csv"))
write.csv(USGS_ElwhaRiverDischarge_During, row.names = FALSE, file =("./Processed/USGS_EwhaRiverDischarge_During_processed.csv"))
```

```{r Two Way Ttest for Water Discharge}
shapiro.test(USGS_ElwhaRiverDischarge_TimeStamp$DailyDischarge[USGS_ElwhaRiverDischarge_TimeStamp$TimeStamp == "After"])
#Results
#W = 0.70006, p-value < 2.2e-16

#shapiro.test(EPAair$Ozone[EPAair$Year == 2019]) EXAMPLE this was run twice, one for each year comparing (2018, 2019)
#therefore mine will be run for the two different time periods i am running my test, before and after indicated by the time stamp column

shapiro.test(USGS_ElwhaRiverDischarge_TimeStamp$DailyDischarge[USGS_ElwhaRiverDischarge_TimeStamp$TimeStamp == "During"])
#results
#W = 0.87227, p-value < 2.2e-16

var.test(USGS_ElwhaRiverDischarge_TimeStamp$DailyDischarge ~ USGS_ElwhaRiverDischarge_TimeStamp$TimeStamp)
#results
#F test to compare two variances
#data:  USGS_ElwhaRiverDischarge_TimeStamp$DailyDischarge by USGS_ElwhaRiverDischarge_TimeStamp$TimeStamp
#F = 2.4947, num df = 765, denom df = 1075, p-value < 2.2e-16
#alternative hypothesis: true ratio of variances is not equal to 1
#95 percent confidence interval:
# 2.189581 2.846533
#sample estimates:
#ratio of variances 
 #         2.494707 

 vartest<-ggplot(USGS_ElwhaRiverDischarge_TimeStamp, aes(x = DailyDischarge, color = TimeStamp)) + geom_freqpoly()
 
 
 DailyDischarge.twosample <- t.test(USGS_ElwhaRiverDischarge_TimeStamp$DailyDischarge ~ USGS_ElwhaRiverDischarge_TimeStamp$TimeStamp) 
 
DailyDischarge.twosample
#Results
#lternative hypothesis: true difference in means is not equal to 0
#95 percent confidence interval:
# -5.269339  2.453161
#sample estimates:
# mean in group After mean in group During 
 #           46.89956             48.30765 


DailyDischarge.twosample$p.value
#results
#0.4744595


#Format as GLM

DailyDischarge.twosample2 <- lm(USGS_ElwhaRiverDischarge_TimeStamp$DailyDischarge ~ USGS_ElwhaRiverDischarge_TimeStamp$TimeStamp) 
summary(DailyDischarge.twosample2)
plot(DailyDischarge.twosample2)


#From the shapiro test, we can see that our data is not normally distributed, so in order to do away with a normal distribution assumption, we need to perform wilcoxon test



```

```{r Wilcoxon test for water discharge}
DailyDischarge.twosample.wilcox <- wilcox.test(USGS_ElwhaRiverDischarge_TimeStamp$DailyDischarge ~ USGS_ElwhaRiverDischarge_TimeStamp$TimeStamp) 

DailyDischarge.twosample.wilcox
#Results
#Wilcoxon rank sum test with continuity correction
#W = 349134, p-value = 2.179e-08
#alternative hypothesis: true location shift is not equal to 0

#Interpretation significant difference in daily water discharge during and after the dam removal process(p value<0.0001, Wilcoxon test)


#Graph this?


```

```{r Two way T-test for Total SEdiment Discharge}

shapiro.test(USGS_ElwhaRiverDischarge_TimeStamp$TotalSedimentDischarge[USGS_ElwhaRiverDischarge_TimeStamp$TimeStamp == "After"])
#W = 0.21518, p-value < 2.2e-16
shapiro.test(USGS_ElwhaRiverDischarge_TimeStamp$TotalSedimentDischarge[USGS_ElwhaRiverDischarge_TimeStamp$TimeStamp == "During"])
#W = 0.37967, p-value < 2.2e-16
#Not normal distribution

#To the wilcoxon test
```

```{r Wilcoxon Test for Total Sediment Discharge}
SedimentDischarge.twosample.wilcox <- wilcox.test(USGS_ElwhaRiverDischarge_TimeStamp$TotalSedimentDischarge ~ USGS_ElwhaRiverDischarge_TimeStamp$TimeStamp) 

SedimentDischarge.twosample.wilcox
#W = 209858, p-value < 2.2e-16
#Interpretation significant difference in daily total sediment discharge during and after the dam removal process(p value<0.0001, Wilcoxon test)
```

```{r Boxplot creation }
ggplot(USGS_ElwhaRiverDischarge_TimeStamp) +
geom_boxplot(aes(x = TimeStamp, y = TotalSedimentDischarge), color="skyblue4", alpha=0.75)+
  xlab("Timeframe in the Dam \nRemoval Process")+
  ylab("Total Daily Sediment \nDischarge (tonnes)")+
  mytheme
#summary stats
tapply(USGS_ElwhaRiverDischarge_TimeStamp$TotalSedimentDischarge, USGS_ElwhaRiverDischarge_TimeStamp$TimeStamp, summary)




ggplot(USGS_ElwhaRiverDischarge_TimeStamp) +
geom_boxplot(aes(x = TimeStamp, y = DailyDischarge), color="skyblue4", alpha=0.75)+
  xlab("Timeframe in the Dam \nRemoval Process")+
   ylab(expression(Daily~Discharge~(m^2/s)))+
  mytheme
#summary stats
tapply(USGS_ElwhaRiverDischarge_TimeStamp$DailyDischarge, USGS_ElwhaRiverDischarge_TimeStamp$TimeStamp, summary)
```


## Question 2: Can we predict sediment discharge from water flow?
>Perform a linear regression- more info from lesson 12 of class lessons

```{r Simple Linear REgression}
sedimentandflowregression<-lm(USGS_ElwhaRiverDischarge_TimeStamp$DailyDischarge~USGS_ElwhaRiverDischarge_TimeStamp$TotalSedimentDischarge)

summary(sedimentandflowregression)

cor.test(USGS_ElwhaRiverDischarge_TimeStamp$DailyDischarge, USGS_ElwhaRiverDischarge_TimeStamp$TotalSedimentDischarge)
#results
#cor 0.7346606  #significant positive correlation


#Let's graph...
par(mfrow = c(2,2), mar=c(1,1,1,1)) 
plot(sedimentandflowregression)
par(mfrow = c(1,1))

dischargebywaterflow <-
ggplot(USGS_ElwhaRiverDischarge_TimeStamp, 
       aes(x = DailyDischarge, y = TotalSedimentDischarge)) + 
geom_point( alpha=0.5)+
  geom_smooth(method = 'lm', se = FALSE, color = "skyblue")+
  xlab(expression(Daily~Water~Discharge~(m^2/s)))+
  ylab("Daily Total Sediment \nDischarge (tonnes)")
print(dischargebywaterflow)
```

#Question 3: Does daily water discharge predict total sediment discharge during and after the two part dam removal process?
a linear regression and graph for during and a linear regression and graph for after
compare the two

```{r Linear Regression During}
duringregression<-lm(USGS_ElwhaRiverDischarge_During$DailyDischarge~USGS_ElwhaRiverDischarge_During$TotalSedimentDischarge)
summary(duringregression)

cor.test(USGS_ElwhaRiverDischarge_During$DailyDischarge, USGS_ElwhaRiverDischarge_During$TotalSedimentDischarge)
#cor 0.7038759 

#graph
dischargebywaterflow_during <-
ggplot(USGS_ElwhaRiverDischarge_During, 
       aes(x = DailyDischarge, y = TotalSedimentDischarge)) + 
geom_point( alpha=0.5)+
  geom_smooth(method = 'lm', se = FALSE, color = "skyblue")+
  xlab(expression(Daily~Water~Discharge~(m^2/s)))+
  ylab("Daily Total Sediment \nDischarge (tonnes)")
print(dischargebywaterflow_during)

```


```{r linear regression After}
afterregression<-lm(USGS_ElwhaRiverDischarge_After$DailyDischarge~USGS_ElwhaRiverDischarge_After$TotalSedimentDischarge)
summary(afterregression)

cor.test(USGS_ElwhaRiverDischarge_After$DailyDischarge, USGS_ElwhaRiverDischarge_After$TotalSedimentDischarge)
#cor 0.7706031 

#graph
dischargebywaterflow_after <-
ggplot(USGS_ElwhaRiverDischarge_After, 
       aes(x = DailyDischarge, y = TotalSedimentDischarge)) + 
geom_point( alpha=0.5)+
  geom_smooth(method = 'lm', se = FALSE, color = "skyblue")+
  xlab(expression(Daily~Water~Discharge~(m^2/s)))+
  ylab("Daily Total Sediment \nDischarge (tonnes)")
print(dischargebywaterflow_after)


```

